---
title: "thema09_logboek"
author: "Isabella Hofstede"
date: '2023-09-13'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(dplyr)
library(grid)
```

## week 1

# Context:
This week a dataset will be chosen with the aim of creating a supervised ML model for classification. The choice was made to pick the a dataset on bird classification by bonestructure: https://www.kaggle.com/datasets/zhangjuefei/birds-bones-and-living-habits

There are many kinds of birds. Some are good at flying, others can't fly but run fast. Some swim under water, others go wading in shallow pool. According to their living environments and living habits, birds are classified into different ecological groups. There are 8 ecological groups of birds:

Swimming Birds
Wading Birds
Terrestrial Birds
Raptors
Scansorial Birds
Singing Birds
Cursorial Birds (not included in dataset)
Marine Birds (not included in dataset)
The first 6 groups are covered by this dataset.

Apparently, birds belonging to different ecological groups have different appearances: flying birds have strong wings and wading birds have long legs. Their living habits are somewhat reflected in their bones' shapes. As data scientists we may think of examining the underlying relationship between sizes of bones and ecological groups , and recognizing birds' ecological groups by their bone structure.

# Status of data
The dataset is made up of bird skeletons from collections of the Natural History Museum of Los Angeles County (data provided by Dr. D. Liu of Beijing Museum of Natural History). The dataset is a 420x10 size continuous values unbalanced multi-class dataset. There are 420 birds contained in this dataset. Each bird is represented by 10 measurements (features):

Length and Diameter of Humerus
Length and Diameter of Ulna
Length and Diameter of Femur
Length and Diameter of Tibiotarsus
Length and Diameter of Tarsometatarsus

The class attribute is "type", referring to the ecological group. The bird skeletons belong to 21 orders, 153 genera, 245 species. Each bird has a label for its ecological group:

SW: Swimming Birds
W: Wading Birds
T: Terrestrial Birds
R: Raptors
P: Scansorial Birds
SO: Singing Birds

Pictured below are the attributes with their range and distribution:
```{r}
#load the data
bird_data <- read.table('bird.csv', sep=",", header = 1)
summary(bird_data)
```
# Missing data
The summary shows that there are missing values present. The significance of these entries must be checked. 
 
```{r}
#checking for NA values 
missing_data <- sapply(bird_data, function(x) sum(is.na(x)))
missing_data_id <- bird_data$id[apply(is.na(bird_data), 1, any)]
missing_data
missing_data_id
```
There are a total of 420 entries with 7 having missing values. The missing values are the measurements of certain bones but none of the main attribute data; "type", is missing. This means a choice must be made to discard the missing data or perform imputation, which means replacing the missing data with substitutions. However 7 out of 420 means the instances with missing values account for 2% of the dataset, so then they may just be insignificant enough to remove entirely. Let's also think about why the data is missing; is it significant to specific types of birds or is the data quite random?

```{r}
#checking which types of birds have missing data
missing_data_type <- bird_data$type[apply(is.na(bird_data), 1, any)]
missing_data_type
```
Apparently there are multiple songbirds with "missing bones". These missing values are most likely field data that was not properly formatted into the csv file. We shall further investigate the data to see if removing the entries with mission values will have an impact on the usability of the dataset. 

# Class distribution 
We need to check whether or not the data is equally distributed between the types of birds. 
```{r}
#check the distribution of the bird types
class_distribution <- table(bird_data$type)
class_distribution
```
On a glance we can already see that the distribution follows a curve. A histogram will show the severity of the differences: 

```{r}
#create a histogram per species
ggplot(bird_data, aes(x = type)) + geom_bar(fill = "black") + labs(title = "distribution of bird species", x = "species", y = "count")
```

Since the SO (songbirds) are over represented in the data, it seems fine to remove the entries with missing values. 

# Variation and distributions 
Boxplot:
```{r}

```

Histogram:
```{r}

```




# Correlation 



# Research Question 
Based on the exploratory data analysis the research question will be: Can a machine learning model accurately classify birds into ecological groups based on their bone measurements, and if so, which bone measurements are most informative for classification? 

Which then follows that the null hypothesis will be: There is no significant relationship between bird bone measurements and their ecological groups, and bone measurements are not informative for accurate classification. 

And so the alternative hypothesis: There is a significant relationship between bird bone measurements and their ecological groups, and bone measurements are informative for accurate classification. 